---
title: "R Tutorial 10"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include = FALSE}
library(learnr)
library(tidyverse)
library(RCurl)
library(DescTools)
library(epiR)
library(ResourceSelection)
knitr::opts_chunk$set(echo = FALSE)

dat_url <- getURL("https://raw.githubusercontent.com/aciarleglio/App_Cat/main/BURN13M.csv")
burn_data13 <- tibble(read.csv(text = dat_url))
burn_data11 <- 
  burn_data13 %>% 
  filter(PAIRID %in% c(1,4))

set.seed(1234)
ids <- sample(1:124)
subj <- rep(ids, 2)
undilated <- c(c(rep(1, 8), rep(1, 72), rep(0, 25), rep(0,19)),
                 c(rep(1, 8), rep(0, 72), rep(1, 25), rep(0,19)))
treatment <- factor(c(rep("trt", 124), rep("placebo", 124)), levels = c("placebo", "trt"))
ophtho_data <- tibble(subj, treatment, undilated)
ophtho_data <- ophtho_data[order(ophtho_data$subj),]
pa_tab <- as.table(matrix(c(8,72,25,19), nrow = 2, byrow = TRUE))
rownames(pa_tab) <- colnames(pa_tab) <- c("Normal", "Dilated")
names(dimnames(pa_tab)) <- c("Treated", "Placebo") # list(c("Treated", "Placebo"))

# table(ophtho_data$treatment, ophtho_data$undilated, ophtho_data$subj)

# ophtho_data_for_pa <- 
#   ophtho_data %>% 
#   spread(key = treatment, value = undilated)
# table(ophtho_data_for_pa$trt, ophtho_data_for_pa$placebo)[c(2,1), c(2,1)]
# 
# burn_data11_pa <-
#   burn_data11 %>% 
#   spread(key = DEATH, value = INH_INJ)
# table(burn_data11_pa$`0`, burn_data11_pa$`1`)
# 
# burn_data11_fpat_case <- 
#   burn_data11 %>% 
#   filter(DEATH == 1) %>% 
#   select(PAIR, INH_INJ)
# 
# burn_data11_fpat_cont <- 
#   burn_data11 %>% 
#   filter(DEATH == 0) %>% 
#   select(PAIR, INH_INJ)
# 
# burn_data11_fpat <- full_join(burn_data11_fpat_case, 
#                               burn_data11_fpat_cont, by = "PAIR")

```


## 1 Introduction

In this tutorial we will cover

- How to compute McNemar's test statistic and p-value.
- How to conduct a Mantel-Haenszel analysis of matched data. 
- How to fit conditional logistic regression models.  

We will be using data from the ophthalmology study discussed in Lecture 10.  You can assume that this data set has been stored as a tibble object called ``ophtho_data``.  The variables that we will use are:

|Variable        |Levels                                         | Variable Class in R|
|:--------------:|:---------------------------------------------:|:------------------:|   
| ``subj``       |indicator of eye pairs                         | numeric            |
| ``treatment``  |treatment or placebo (ref. level)              | factor             |
| ``undilated``  |1 = normal, 0 = dilated                        | numeric            |


We will also be using data from a matched-pairs case-control study of burn victims discussed in Lecture 10.  In the lecture, we considered a 1:3 matched data set.  This data set has been reduced to include only one case and one matched control.  Cases (those who died of their injuries) were matched with controls (those who did not die), based on age.  The table below defines the relevant variables.

|Variable        |Levels                                         | Variable Class in R|
|:--------------:|:---------------------------------------------:|:------------------:|   
| ``DEATH``      |1 if died, 0 otherwise                         | numeric            |
| ``TBSA``       |total body surface area of burn                | factor             |
| ``GENDER``     |1 if male, 0 otherwise                         | numeric            |
| ``RACEC``      |1 if white, 0 otherwise                        | numeric            |
| ``INH_INJ``   |1 if inhalation injury involved, 0 otherwise    | numeric            |
| ``FLAME``      |1 if flame involved in burn injury, 0 otherwise| numeric            |
| ``PAIR``       |matched set identifier                         | numeric            |

\begin{table}
\footnotesize
\begin{tabular}{ll}
Variable & Coding \\
\hline
$\mathtt{DEATH}$ & 1 if died, 0 otherwise \\
$\mathtt{TBSA}$ & total body surface area of burn\\
$\mathtt{GENDER}$ & 1 if male, 0 otherwise \\
$\mathtt{RACEC}$ & 1 if white, 0 otherwise \\
$\mathtt{INH\_INJ}$ & 1 if inhalation injury involved, 0 otherwise \\
$\mathtt{FLAME}$ & 1 if flame involved in burn injury, 0 otherwise\\
$\mathtt{PAIR}$ & matched set identifier \\
\end{tabular}
\end{table}

## 2 McNemar's Test

McNemar's test is used to assess association between two paired binary variables. The ``mcnemar.test()`` function is used to compute the test statistic ($Z^2$) and corresponding p-value.  The first argument for the function is the population-averaged 2 x 2 table, and the second argument is ``correct = FALSE`` (to obtain the uncorrected test statistic value).  

For the ophthalmology data, the population-averaged 2 x 2 table is:

```{r, echo = FALSE}
pa_tab
```

Recall that the counts in a population-averaged table tell us how many **pairs** of observations fall in that cell.  

When these counts are available, you can simply create the 2 x 2 table in R and use that in the ``mcnemar.test()`` function.

The code chunk below creates the population-averaged table and then computes McNemar's test statistic and p-value for testing the association between treatment and dilation status with the matched pairs.    

```{r mc_01, exercise = TRUE, exercise.lines = 9}
# create the population-averaged table
# note that this puts the Treated Eye responses correspond to the rows
#  and Placebo Eye responses correspond to the columns
eye_pa_tab <- as.table(matrix(c(8,72,25,19), nrow = 2, byrow = TRUE))
rownames(eye_pa_tab) <- colnames(eye_pa_tab) <- c("Normal", "Dilated")

# compute McNemar's test statistic and p-value
mcnemar.test(eye_pa_tab, correct = FALSE)
```

## 3 Mantel-Haenszel Analysis

In Lecture 10, we saw that we can think of matched pairs data as highly stratified data where the strata correspond to the matched pairs.  We can use Mantel-Haenszel methods to analyze these data.  

Consider the ophthalmology data again.  Here are the first few rows in the data set:

```{r mh_01, echo = FALSE}
head(ophtho_data)
```

We can create a stratified table object using the ``table()`` function (this creates a collection of subject-specific tables) and then use the ``epi.2by2()`` function (from the epiR package) to obtain the Mantel-Haenszel OR and its corresponding confidence interval.  We can also obtain McNemar's test statistic and p-value since McNemar's test is a special case of the Mantel-Haenszel test.

Run the code chunk below to conduct the Mantel-Haenszel analysis on the subject-specific tables from the ophthalmology data set.

```{r mh_02, exercise = TRUE, exercise.lines = 15, warning = FALSE}
# create the stratified table object (i.e., the collection 
#   of subject-specific tables)
# note that the first argument is the row variable (treatment), 
#  the second argument is the column variable (dilation status),
#   and the third argument is the matched pair indicator
# be sure that the orientation of the tables is what you want
# here - we want trt in the first row and placebo in the second and
#   we want undilated = 1 in the first column and 0 in the second
eye_ss_tabs <- table(ophtho_data$treatment, 
                     ophtho_data$undilated,
                     ophtho_data$subj)[c(2,1),c(2,1),]

# use the epi.2by2 function to obtain Mantel-Haenszel statistics
epi.2by2(eye_ss_tabs, conf.level = 0.95)
```

In the lecture, we said that we can only use some of the output from the code above:  

(1) ``Odds ratio (crude)`` results give the population-averaged odds ratio estimate; the corresponding confidence interval is NOT correct.  

(2) ``Odds ratio (M-H)`` results give the subject-specific odds ratio estimate; we CAN use the corresponding confidence interval.  

(3) ``Test that M-H adjusted OR = 1`` results are equivalent to the results from McNemar's test.  

(4) ``Attrib risk (crude) *`` results give the population-averaged risk difference estimate; the corresponding confidence interval is NOT correct.  Also, notice that the population-averaged risk difference estimate is the same as the subject-specific risk difference estimate given on the line showing the ``Attrib risk (M-H) *`` results. This follows from an interesting fact that we did not discuss in the lecture: the population-averaged risk difference is always equivalent to the subject-specific risk difference.  Also, note that the confidence intervals for both measures are NOT correct and should not be used.    

You may get warning messages when you apply the ``epi.2by2()`` function to your stratified table object.  This is OK as long as the function ultimately provides estimates and statistics.  There are instances where this will not happen, as illustrated in the next example.  

Now let's do a Mantel-Haenszel analysis with the burn data to assess if there is an association between death (``DEATH``) and whether or not a person suffered an inhalation injury (``INH_INJ``).  

First we set up the stratified table object.  It's natural to think about ``DEATH`` as the column variable and ``INH_INJ`` as the row variable, but if we construct our tables in this way, we will have problems with the ``epi.2by2()`` function.  To use the ``epi.2by2()`` function to perform an Mantel-Haenszel analysis with matched pairs case-control data, we should put the case-control status (``DEATH``) in the rows and the exposure variable status (``INH_INJ``) in the columns.  The code chunk below correctly sets up the stratified table object and runs the Mantel-Haenszel analysis. 

```{r mh_03, exercise = TRUE, exercise.lines = 12, warning = FALSE}
# create the stratified table object with case-control status (DEATH)
#  in the rows and exposure status (INH_INJ) in the columns
# note that we have to adjust the order of the rows and columns to get 
#  the outcome of death (DEATH = 1) in the first row and the exposure of having
# an inhalation injury (INH_INJ = 1) in the first column
burn_ss_tab <- table(burn_data11$DEATH,
                     burn_data11$INH_INJ, 
                     burn_data11$PAIR)[c(2,1), c(2,1),]

# use the epi.2by2 function to obtain Mantel-Haenszel statistics
epi.2by2(burn_ss_tab, conf.level = 0.95)
```

Try switching the row and column variables in the code above and then run the ``epi.2by2()`` function.  You should get an error message.  This is avoided if you follow the advice above.  
 
## 4 Conditional Logistic Regression 

In Lecture 10, we learned that we can use conditional logistic regression to assess association(s) between a binary outcome and one or multiple explanatory variables when the data are paired or clustered.  

We can fit conditional logistic regression models using the ``clogit()`` function from the survival package.  Be sure to install and load this package on your own computer before trying to use the ``clogit()`` function.  You can assume that the package is loaded here.  

The arguments for the ``clogit()`` function are similar the the arguments of the ``glm()`` function.  The arguments for the ``clogit()`` function are:

(1) **Model formula**: ``binary_resp ~ x_1 + x_2 + x_3 + strata(pair_id)``.  This is similar to the model formula entered into the ``glm()`` function when we fit a standard logistic model, BUT you need to include a term that identifies how the observations are paired/clustered.  This is done by adding ``strata(pair_id)`` to the formula, where ``pair_id`` is the variable in the data set that identifies pair/cluster membership.  

(2) **Data frame/tibble**: As with the ``glm()`` function, we need to tell the ``clogit()`` function where the variables in the formula are coming from.  

Notice that we do not have to specify ``family = binomial(link = "logit")`` in the ``clogit()`` function.  That is because this function only fits conditional **logistic** models.  

The code chunk below fits a conditional logistic model to the burn data with ``DEATH`` as the binary response and with the predictors inhalation injury (``INH_INJ``) and total body surface area of burn (``TBSA``).  Recall that the variable for matched pair membership is ``PAIR``.  

```{r cl_01, exercise = TRUE, exercise.lines = 5, warning = FALSE}
# fit the conditional logistic model and store it as clogit object
cl_mod_01 <- clogit(DEATH ~ INH_INJ + TBSA + strata(PAIR), data = burn_data11)
# use the summary function to extract the model estimates and tests
summary(cl_mod_01)
```


## More Practice

**Exercise 1.** A study was done to assess pain level (low/high) before a treatment was administered (measurement 1) and then again 10 days after treatment was administered (measurement 2).  12 subjects had low pain level at both measurements, none had low pain level for the first measurement and high pain level for the second, 13 had high pain level for the first measurement and low pain level for the second, and 37 had high pain level at both measurements.  Compute McNemar's test statistic and corresponding p-value for the test of the null hypothesis of no association between treatment and pain level.           

```{r ex1, exercise = TRUE, exercise.lines = 5, message = FALSE}

```

```{r ex1-solution}
# create the population-averaged table
pain_pa_tab <- as.table(matrix(c(12,0,13,37), nrow = 2, byrow = TRUE))
# obtain McNemar's test statistic and p-value
mcnemar.test(pain_pa_tab, correct = FALSE)
```

**Exercise 2.** Suppose that we want to assess the association between death and gender using the burn data.  Conduct the Mantel-Haenszel analysis for this association.   

```{r ex2, exercise = TRUE, exercise.lines = 7, message = FALSE, warning = FALSE}

```

```{r ex2-solution}
# create the stratified table object
burn_ss_tab2 <- table(burn_data11$DEATH,
                     burn_data11$GENDER, 
                     burn_data11$PAIR)[c(2,1), c(2,1),]
# use the epi.2by2 function to obtain Mantel-Haenszel statistics
epi.2by2(burn_ss_tab2, conf.level = 0.95)
```

**Exercise 3.** Fit a condition logistic regression model to assess the association between death and gender using the burn data.

```{r ex3, exercise = TRUE, exercise.lines = 7, message = FALSE}

```

```{r ex3-solution}
# fit the conditional logistic model and store it as clogit object
cl_mod_02 <- clogit(DEATH ~ GENDER + strata(PAIR), data = burn_data11)
# use the summary function to extract the model estimates and tests
summary(cl_mod_02)
```