---
title: "R Tutorial 08"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include = FALSE}
library(learnr)
library(tidyverse)
library(RCurl)
library(gmodels)
library(epiR)
library(Exact)
library(DescTools)
library(lmtest)
library(emmeans)
library(Stat2Data)
knitr::opts_chunk$set(echo = FALSE)

set.seed(12345)
age <- round(runif(n = 1500, min = 0, max = 95), 2)
lp <- 0.0011*(age - 48)^2
pr <- exp(lp) / (1 + exp(lp))
death <- rbinom(n = 1500, size = 1, prob = pr)
age_death_data <- data.frame(age, death)

set.seed(123)
weight <- runif(1000, min = 120, max = 250)
diabetes_lp <-  -4.5 + 0.0006*(weight - 130)^2 
diabetes_pr <- exp(diabetes_lp) / (1 + exp(diabetes_lp))
diabetes <- rbinom(1000, size = 1, prob = diabetes_pr)
diabetes_data <- tibble(diabetes, weight)

```


## 1 Introduction

In this tutorial we will cover

- How to construct an empirical logit plot for a numerical predictor.
- How to create centered and standardized variables.

We will be using simulated data from a hypothetical study of age and risk of death from Lecture 08.  You can assume that this data set has been stored as a tibble object called ``age_death_data``.  The variables that we will use are:

|Variable           |Levels                             | Variable Class in R|
|:-----------------:|:---------------------------------:|:------------------:|   
| ``death``         |1 = dead, 0 = alive                | numeric            |
| ``age``           |age in years                       | numeric            |

## 2 Empirical Logit Plots

An empirical logit plot is easy to construct using the ``emplogitplot1()`` function from the Stat2Data package.  Be sure to install and load the Stat2Data package on your computer before you try to use the ``emplogitplot1()`` function.  

The ``emplogitplot1()`` function takes three arguments:  

(1) A formula specifying the binary response and the numerical predictor.  If $Y$ is the binary response and $X$ is the numerical predictor then the formula argument should be written as ``Y ~ X``.  

(2) ``data``: The data frame/tibble object that contains the response and predictor variables.  

(3) ``ngroups``: The number of intervals that the numerical predictor should be split into.  If you select, for example, ``ngroups = 5``, then the ``emplogitplot1()`` function will categorize the numerical predictor into 5 levels (intervals) with approximately the same number of observations in each level (interval), (i.e., 20\% of the observations fall in each level when ``ngroups = 5``).      

The code below constructs an empirical logit plot of age vs. the log odds of death using data in the ``age_death_data`` tibble object. It categorizes the age variable into 10 intervals.  Run the code to obtain the empirical logit plot.  

```{r elp01, exercise = TRUE, exercise.lines = 4}
emplogitplot1(death ~ age, data = age_death_data, ngroups = 10)
```

## 3 Centering and Standardizing

We can create centered or standardized variables in R by combining piping and the mutate function.  

Centering a numerical variable means subtracting its sample mean from all observations of that variable.  

The code below creates a centered age variable called ``age_cent`` based on the ``age`` variable in the ``age_death_data`` tibble object.  It stores the new ``age_cent`` variable in the ``age_death_data`` tibble object.  Run the code below to create the centered age variable and use the ``summary()`` function to inspect the ``age_cent`` variable.      

```{r cent01, exercise = TRUE, exercise.lines = 6}
age_death_data <- 
  age_death_data %>% 
  mutate(age_cent = age - mean(age))
summary(age_death_data$age_cent)
```

Standardizing a numerical variable means subtracting its sample mean from all observations of that variable and then dividing the results by the sample standard deviation.  

The code below creates a standardized age variable called ``age_stdz`` based on the ``age`` variable in the ``age_death_data`` tibble object.  It stores the new ``age_stdz`` variable in the ``age_death_data`` tibble object.  Run the code below to create the standardized age variable and use the ``summary()`` function to inspect the ``age_stdz`` variable.

```{r stdz01, exercise = TRUE, exercise.lines = 6}
age_death_data <- 
  age_death_data %>% 
  mutate(age_stdz = (age - mean(age)) / sd(age))
summary(age_death_data$age_stdz)
```

## More Practice

Data from a hypothetical study on the relationship between weight and diabetes in men is provided in the tibble object called ``diabetes_data``.  The variables that we will use are:

|Variable           |Levels                             | Variable Class in R|
|:-----------------:|:---------------------------------:|:------------------:|   
| ``diabetes``      |1 = diabetes, 0 = no diabetes       | numeric            |
| ``weight``        |weight in pounds                   | numeric            |

**Exercise 1.** Construct an empirical logit plot to determine if there is a linear association between the log odds of diabetes and weight.  Split the ``weight`` variable into 10 intervals with approximately equal numbers of observations falling in each interval.   

```{r ex1, exercise = TRUE, exercise.lines = 7, message = FALSE}

```

```{r ex1-solution}
emplogitplot1(diabetes ~ weight, data = diabetes_data, ngroups = 10)
```

**Exercise 2.** Create a centered version of the ``weight`` variable called ``weight_cent`` and store this in the ``diabetes_data`` tibble.

```{r ex2, exercise = TRUE, exercise.lines = 7, message = FALSE}

```

```{r ex2-solution}
diabetes_data <- 
  diabetes_data %>% 
  mutate(weight_cent = weight - mean(weight))
```

**Exercise 3.** Create a standardized version of the ``weight`` variable called ``weight_stdz`` and store this in the ``diabetes_data`` tibble.

```{r ex3, exercise = TRUE, exercise.lines = 7, message = FALSE}

```

```{r ex3-solution}
diabetes_data <- 
  diabetes_data %>% 
  mutate(weight_stdz = (weight - mean(weight)) / sd(weight))
```